*Salmonella* Environmental Persistence Informs Management Relevant to Avian and Public Health

### 2023-04-09 3. Analysis Script/StatisticalAnalysis.qmd
Revamp: Analysis [IP]

glimpse (FD_Data)
```

# **Data Fitting/Statistical Analysis**
Let's Begin with some Linear Model Fitting; I will try this multiple ways, with several predictors.
```{r}
lmog<-linear_reg() %>% 
  set_engine("lm")

@@

TableTidy_PTD<- here::here("4. Results","PTD_TableTidy.png")

ggsave(lmfit_PT, file=TableTidy_PTD)

@@

PlotSum_PD1= here("results", "PD_results_PlotSum.rds") #KP Double check path going to MADA Project not folder
saveRDS(lmall_PT, file = "PlotSum1_PD")

@@

plot_summs(lmall_PT, robust=TRUE) #KP NOTE: I need to clean this up so it is a bit more readable, must fullscreen it

PlotSum_PD1= here("results", "PD_results_PlotSum.rds") #KP Double check path going to MADA Project not folder
saveRDS(lmall_PT, file = "PlotSum1_PD")

#### Model fit another way
# fit linear model using Salmonella_Pos as outcome, Table as predictor
lmfit3.0 <- lm(Salmonella_Positive~Table, PT_Data) 
lmfit3.0
plot(lmfit3.0)

@@

# save fit results table  
Table1_PD= here("results", "PD_results_table.rds") #KP Double check path going to MADA Project not folder
saveRDS(lmtable1, file = "Table1_PD")

@@

#Results
print(lmtable2)

#Save  
Table2_PD = here("results", "PD_results_table2.rds")
saveRDS(lmtable2, file = "Table2_PD")

############################
#### Second model fit

#Doing the same for FEEDER DATA
lmog_FD<-linear_reg() %>% 
  set_engine("lm")

@@

TableTidy_FD= here("results", "FD_results_TableTidy.rds") #KP Double check path going to MADA Project not folder
saveRDS(lmfit_FD, file = "TableTidy_FD")

@@

PlotSum_FD1= here("results", "FD_results_PlotSum.rds") #KP Double check path going to MADA Project not folder
saveRDS(lmfit_FD, file = "PlotSum1_FD")

@@

TableTidy_FD2= here("results", "FD_results_TableTidy2.rds") #KP Double check path going to MADA Project not folder
saveRDS(lmfit_FD2, file = "TableTidy_FD2")

@@

PlotSum_FD2= here("results", "FD2_results_PlotSum.rds") #KP Double check path going to MADA Project not folder
saveRDS(lmfit_FD2, file = "PlotSum2_FD")

@@

plot_summs(lmall_FD, robust=TRUE) #KP NOTE: I need to clean this up so it is a bit more readable, must fullscreen it

PlotSum_FDALL= here("results", "FD_results_PlotSumALL.rds") #KP Double check path going to MADA Project not folder
saveRDS(lmall_FD, file = "PlotSumALL_FD")

#Let's Start with Defining our model 
#Because both out data is continuous, I am going to use a linear regression model

#After the LR, I will create my recipe, create my workflow, create model using the workflow

#And finally try the model on data
```

>>>>

glimpse (FD_Data)
```
# 2. **Data Fitting/Statistical Analysis: Picnic Table**
Let's Begin with some Linear Model Fitting; I will try this multiple ways, with several predictors.
```{r}
lmog<-linear_reg() %>% 
  set_engine("lm")

@@

@@

@@

plot_summs(lmall_PT, robust=TRUE) #KP NOTE: I need to clean this up so it is a bit more readable, must fullscreen it
```

```{r} 
#### Model fit another way
# fit linear model using Salmonella_Pos as outcome, Table as predictor
lmfit3.0 <- lm(Salmonella_Positive~Table, PT_Data) 
lmfit3.0
plot(lmfit3.0)

@@

@@

#Results
print(lmtable2)
```

Second model fit
```{r}
#Doing the same for FEEDER DATA
lmog_FD<-linear_reg() %>% 
  set_engine("lm")

@@

@@

@@

@@

@@

plot_summs(lmall_FD, robust=TRUE) #KP NOTE: I need to clean this up so it is a bit more readable, must fullscreen it
```
# **IP Analysis KP finish by 4/9!!**

#**Time to event**
```{r}
library(finalfit)
library(forcats)

ff_glimpse(PT_Data)

PT_Data= PT_Data %>%
  mutate(

  )
```

#**Cox regression**
```{r}
```

### 2023-04-09 5. Manuscript/Manuscript.qmd
Revamp: Analysis [IP]

---
title: "*Salmonella* Environmental Persistence Informs Management Relevant to Avian and Public Health"
subtitle: ""
author: Kimberly M. Perez
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: Ref.bib
---

@@

Given the uncertainty of *Salmonella's* persistence on various surface materials, from my data I would like to answer: - What is the prevalence of *Salmonella* over time? - How long does *Salmonella* persist on different surface materials (e.g., wood, plastic, and those with an antimicrobial coating) exposed to various climatic factors (e.g., rain, sun, changing temperatures)

From these questions, I hope to better inform the public and public health practitioners on safer feeder handling methods and precautions to take when frequenting public locations where humans and wildlife (e.g., ibis) interact.

@@

The ibis picnic table trial was conducted at Dreher Park, a public park in Palm Beach County frequented by white ibis who are known to carry *Salmonella*. Because humans and ibis both utilize the wooden picnic tables at the park, three tables under a single pavilion with similar environmental conditions (e.g., exposure to sunlight, rainfall, etc.) were selected. All tables were divided into four quadrants (e.g., tabletop plus bench) and ibis were encouraged to feed on the picnic tables until they deficated. Pictures were taken of every quadrant daily to facilitate the counting of new fecal piles. If feces were not deposited in a quadrant, fresh feces from the concrete pad was collected and placed onto the quadrant. To determine persistence, one large fecal pile from each quadrant was selected as the "persistence pile" and monitored and sampled daily for 8 days. To determine table prevalence, a pooled sample was taken daily from every quadrant by swabbing the surface picnic table surface with a sampling sponge, avoiding the designated "persistence pile". To maintain cleanliness for other park patrons, new piles were not smeared with the sampling sponge. Samples were placed into 20 mL of dulcitol selenite and maintained at room temperature prior to being shipped to the University of Georgia (UGA), every 2-4 days, for processing.

@@

@@

Before attention can be focused on the data analysis portion, removal of unnecessary variables and visualizing data will provide a clearer picture of the data. This section will focus on data cleaning and wrangling. First, both datasets must be loaded. Prior to loading any data into RStudio, ensure that a new Quarto or RMarkdown File is established and linked to the preferred working directory. Basic packages need to be loaded to read in the data. Such packages to download include read_csv(), read_excel(), and here(), each of which can be loaded by utilizing the library() function (e.g., library(here)). Next, the datasets from both the Ibis and Feeder trials are loaded into RStudio utilizing the appropriate function (e.g., read_csv() for .csv files and read_excel() for .xlsx files). After data is loaded into R, it is best to get acquainted with the data by utilizing basic functions such as glimpse(). Glimpse provides me with insight into my data such as column names, number of rows and how each row is classified (e.g., character or numeric). After viewing the data I can work on cleaning the data by removing unnecessary variables. I would like to make the data a little more uniform and easier to work with. With that, I will change "Yes" to 1, and *NAs* to 0. I believe this will help me further in the analysis process. Given the variablility in the column names, I will then clean the names up to make them easier to include in my code as I move forward. With all of that, it is finally time to visualize the data by making some tables and/or graphs.

Here, some more packages were loaded that I forsaw using in this section. Before visualization began, the glimpse() function was utilized to view the recently cleaned data. Given the uncertainty of *Salmonella's* persistence on various surface materials and the information these trials hoped to provide, the summary() function was utilized to provide a quick glimpse of basic result summaries for positive *Salmonella* samples and if the pile present was the original pile or not. Next, I continued to explore these two variables of interest by plotting each using a stacked bar graph via ggplot().

I followed the same workflow for the Feeder Data, however, this dataset required less wrangling and did not provide as many exploratory options as the main focus was on feeder type, date, and the presence/absence of *Salmonella* for each day the feeders were sampled.

@@

# **Future Project PARTS for KP to populate**

@@

Of 96 deposited feces, we obtained 27% positive.
*Salmonella* persisted for an average of 3 days, and maximum of 4 days on
wooden picnic tables.

## *Feeder Trial*
We confirmed successful inoculation of all
feeders with *Salmonella* (presence on day post inoculation \[DPI\]
0). Further, results indicate that *Salmonella* may persist on
some feeders up to DPI 5 and current analysis indicates that *Salmonella* persisted
on both coated and uncoated plastic feeders at a higher rate than wood and wood
coated. Figure 2 depicts the percent of samples positive based on feeder type
throughout the study period. While plastic and plastic coated feeders had an
overall higher percent positivity, a spike in percent positivity in wood feeders occurred on DPI 5 after a rain event. Further analyses should be conducted to determine whether environmental
conditions (e.g., rain event) had a significant influence on persistence.

Our samples persisted for an average of 2 days (feeders), 3 days (picnic tables) with 4 (picnic tables)
and 5 (feeders) being the maximum number of days persisted. The varied
persistence during both trials could be as a result of a lack of moisture, the
fluctuation in ambient temperature during sampling, and water activity. Our studies
suggest that *Salmonella* can have prolonged persistence on shared spaces,
persistence depends on surface material, and precautionary measures (i.e.,
increased hygiene) should be adopted to reduce human exposure.

>>>>

---
title: "*Salmonella* Environmental Persistence Informs Management Relevant to Avian and Public Health"
subtitle: ""
author: Kimberly M. Perez
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: Ref.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Given the uncertainty of *Salmonella's* persistence on various surface
materials, questions such as:

-   What is the prevalence of *Salmonella* over time?

-   How long does *Salmonella* persist on different surface materials
    (e.g., wood, plastic, and those with an antimicrobial coating)?

-   Is there a greater risk of acquiring *Salmonella* from objects that
    both humans and animals use (e.g. picnic tables in parks)?

From these questions, we hope to better inform the public and public
health practitioners on safer feeder handling methods and precautions to
take when frequenting public locations where humans and wildlife (e.g.,
ibis) interact.

@@

The ibis picnic table trial was conducted at Dreher Park, a public park
in Palm Beach County frequented by white ibis who are known to carry
*Salmonella*. Because humans and ibis both utilize the wooden picnic
tables at the park, three tables under a single pavilion with similar
environmental conditions (e.g., exposure to sunlight, rainfall, etc.)
were selected. All tables were divided into four quadrants (e.g.,
tabletop plus bench) and ibis were encouraged to feed on the picnic
tables until they deficated. Pictures were taken of every quadrant daily
to facilitate the counting of new fecal piles. If feces were not
deposited in a quadrant, fresh feces from the concrete pad was collected
and placed onto the quadrant. To determine persistence, one large fecal
pile from each quadrant was selected as the "persistence pile" and
monitored and sampled daily for 8 days. To determine table prevalence, a
pooled sample was taken daily from every quadrant by swabbing the picnic table surface with a sampling sponge, avoiding the designated "persistence pile". To maintain cleanliness for other park
patrons, new piles were not smeared with the sampling sponge. Samples
were placed into 20 mL of dulcitol selenite and maintained at room
temperature prior to being shipped to the University of Georgia (UGA),
every 2-4 days, for processing.

@@

The data from the Picnic Table Trial in South Florida was collected by an
undergraduate student of Dr. Sonia M. Hernandez and provided to me by Dr.
Hernandez. All data from the Feeder Trial was collected by me, Kimberly
Perez under the guidance of Dr. Sonia Hernandez.

@@

Before attention can be focused on the data analysis portion, removal of
unnecessary variables and visualizing data will provide a clearer
picture of the data. This section will focus on data cleaning and
wrangling. First, both datasets must be loaded. Prior to loading any
data into RStudio, ensure that a new Quarto or RMarkdown File is
established and linked to the preferred working directory. Basic
packages need to be loaded to read in the data. Such packages to
download include read_csv(), read_excel(), and here(), each of which can
be loaded by utilizing the library() function (e.g., library(here)).
Next, the datasets from both the Ibis and Feeder trials are loaded into
RStudio utilizing the appropriate function (e.g., read_csv() for .csv
files and read_excel() for .xlsx files). After data is loaded into R, it
is best to get acquainted with the data by utilizing basic functions
such as glimpse(). Glimpse provides users with insight into the data
such as column names, number of rows and how each row is classified
(e.g., character or numeric). After viewing the data, cleaning the data
by removing unnecessary variables can occur.

Ensuring the data is uniform by removing unnecessary variables and renaming columns will assist in the analysis process downstream. To begin, "Yes" responses should be coded to produce a 1, while *NAs* would produce a 0. Given the variability in the column names, standardizing or renaming them a preferred name I will streamline future coding or analysis. With all of that, it is finally time to visualize the data by making some tables and/or graphs.

More packages were loaded that would assist in the visualization process before begining this process. Before visualization began, the glimpse() function was utilized
to view the recently cleaned data. Given the uncertainty of
*Salmonella's* persistence on various surface materials and the
information these trials hoped to provide, the summary() function was
utilized to provide a quick glimpse of basic result summaries for
positive *Salmonella* samples and if the pile present was the original
pile or not. Exploration continued on these two variables of
interest by plotting each using a stacked bar graph via ggplot().

The same workflow for the Feeder Data was utilized, however, this dataset
required less wrangling and did not provide as many exploratory options
as the main focus was on feeder type, date, and the presence/absence of
*Salmonella* for each day the feeders were sampled.

@@

@@

### **Persistence**
On Day 1 of the study, four fecal piles per table (one per quadrant, n = 12) were designated for tracking *Salmonella* persistence.  **Salmonella** was isolated from three of the 12 designated persistence study piles on Day 1, allowing the persistence to be tracked on these three piles.   Persistence, expressed as the number of days that a positive fecal pile on Day 1 remained positive, ranged from 1 to 4 days with an average persistence of 3 days. The probability of an already positive sample to remain positive the next day declined consistently throughout the study period.

### **Prevalence**
Of 96 deposited feces, we obtained 27% positive. *Salmonella* persisted for an average of 3 days, and maximum of 4 days on wooden picnic tables. *Salmonella* was isolated from all tables, at least once. While there appears to be a significant positive correlation between the number of new fecal piles and the probability of yielding a positive pooled sample. The probabilities of a positive sample varied throughout the week, with day 4 having the lowest probability of positive *Salmonella* isolation and day 7 having the most error.

## **Feeder Trial**
We confirmed successful inoculation of all feeders with *Salmonella*
(presence on day post inoculation \[DPI\] 0). Further, results indicate
that *Salmonella* may persist on some feeders up to DPI 5 and current
analysis indicates that *Salmonella* persisted on both coated and
uncoated plastic feeders at a higher rate than wood and wood coated.
Figure 2 depicts the percent of samples positive based on feeder type
throughout the study period. While plastic and plastic coated feeders
had an overall higher percent positivity, a spike in percent positivity
in wood feeders occurred on DPI 5 after a rain event. Further analyses
should be conducted to determine whether environmental conditions (e.g.,
rain event) had a significant influence on persistence.

Our samples persisted for an average of 2 days (feeders), 3 days (picnic
tables) with 4 (picnic tables) and 5 (feeders) being the maximum number
of days persisted. The varied persistence during both trials could be
a result of a lack of moisture, the fluctuation in ambient temperature
during sampling, and water activity. Our studies suggest that
*Salmonella* can have prolonged persistence on shared spaces,
persistence depends on surface material, and precautionary measures
(i.e., increased hygiene) should be adopted to reduce human exposure.
